Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-05-02 manuscript/manuscript.Rmd
Head reset issues

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)

@@

filled_file_metadata <- read_csv("../metadata/filled_file_metadata.csv") %>%
  filter(polarity=="pos") %>%
  select(-polarity)
clean_stans <- read_csv("../metadata/clean_stans.csv")
untarg_feats <- read_csv("../untargeted/all_feats.csv")
untarg_peaks <- read_csv("../untargeted/all_peaks.csv") %>%
  filter(str_detect(filename, "_Smp_"))
targ_peaks <- read_csv("../targeted/all_peaks.csv") %>%
    mutate(filename=paste0(filename, ".mzML"))

@@

Mesoscale eddies significantly alter open ocean environments such as the North Pacific Subtropical Gyre. Previous studies have explored eddy effects on biogeochemical measurements and community composition but not on the molecular composition of particulate organic matter. This study reports the exact concentration of 67 metabolites and relative abundances for 640 molecular features to understand how mesoscale eddies impact the NPSG's metabolome during two cruises in 2017 and 2018. We find that many metabolites track biomass trends, but biosignals like isethionic acid, homarine, and trigonelline linked to picoeukaryotes were enriched at the deep chlorophyll maximum of the cyclonic features, while degradation products such as arsenobetaine were enriched in anticyclones. Additionally, we identify taurine betaine (N,N,N-trimethyltaurine) in open ocean samples for the first time. By analyzing depth variability (accounting for 20-40% of metabolomic variability across ~150 meters) and isopycnal disruption (explaining 10-20% of variability across a sea level anomaly range of 40 centimeters), this analysis constrains the importance of mesoscale eddies in modeling the chemical composition of particulate matter in the largest biomes on the planet.

@@

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016; @Rii2022]. These mesoscale features introduce nutrients into the euphotic zone and alter the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

Here, we use measurements of metabolites in particles across an eddy dipole in the North Pacific Subtropical Gyre (NPSG) to categorize both known compounds and unidentified mass features based on their co-occurrence with a set of simultaneously measured physical, chemical, and biological parameters. The gyre in particular represents an intriguing opportunity for metabolomic research because it has been shown to contain the highest proportion of chemical features with unknown structure and function, many of which have highly variable responses to changes in latitude, depth, and across diel cycles [@Heal2021; @Boysen2021; @Kumler2023]. This study connects biogeochemical measurements to a molecular-scale perspective of particulate organic matter, linking the chemical currencies that fuel metabolism and interactions among marine microbes to the mesoscale environment in which the currencies are found.

# Results
## Metabolome variability across the eddy dipole

We found that the metabolome clearly differed between the cyclonic and anticyclonic samples when all molecular features were analyzed. Multivariate statistics detected the largest differences in particulate matter composition at the DCM and 175 meter samples (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by sea level anomaly (SLA) alone (PERMANOVA, p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the 175m samples, 12% of the variance was explained by SLA (PERMANOVA, p = 0.008) and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to largely capture variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (PERMANOVA, p=0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize the difference between eddies in a lower-dimensional space, we plotted an "average" metabolite peak area per station at each depth by taking the median value of each of the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was more abundant than in other samples at this same depth. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

```{r multivar_analysis object creation - nmds/pca/permanova}
set.seed(123)
multivar_analysis <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  filter(ms_run=="MT") %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(feature, filename, norm_area) %>%
      group_by(feature) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      mutate(norm_area=scale(norm_area)[,1]) %>%
      pivot_wider(names_from = "feature", values_from = norm_area) %>%
      column_to_rownames("filename") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE) %>%
    vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(filename=rownames(metab_mat)) %>%
      left_join(filled_file_metadata %>% select(filename, sla_class), by="filename")
    vegan::adonis2(
      metab_mat~sla_class, data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(perm_r2=map_dbl(perm_output, pluck, "R2", 1)) %>%
  mutate(perm_pval=map_dbl(perm_output, pluck, "Pr(>F)", 1)) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(nmds_stress=paste("s =", round(nmds_stress, digits = 3))) %>%
  mutate(perm_r2=round(perm_r2, digits = 2)) %>%
  mutate(perm_pval=round(perm_pval, digits = 4)) %>%
  mutate(perm_label=paste0("R2 = ", perm_r2, "\np = ", perm_pval))
```

```{r figure nmds and median metabolite across eddy transect}
nmds_sla_gp <- multivar_analysis %>%
  mutate(nmds_points=map(nmds_output, function(nmds_output_i){
    nmds_output_i %>% pluck("points") %>% as.data.frame()
  })) %>%
  select(depth, nmds_points) %>%
  unnest_longer(nmds_points, indices_to = "filename") %>%
  unnest_wider(nmds_points) %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_corr)) %>%
  mutate(MDS2=ifelse(str_detect(depth,"Deep"), MDS2*-1, MDS2)) %>%
  ggplot() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, fill=sla_corr), color="black", size=4, pch=21) +
  geom_label(x=-Inf, y=-Inf, aes(label=nmds_stress), hjust=0, vjust=0, 
             data = multivar_analysis, label.r = unit(0, "in")) +
  geom_label(x=-Inf, y=Inf, aes(label=perm_label), hjust=0, vjust=1,
             data = multivar_analysis, label.r = unit(0, "in"), lineheight = .7) +
  facet_wrap(~depth, nrow=1) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"),
  ) +
  labs(fill="Sea level\nanomaly\n(cm)\n") +
  theme_bw() +
  coord_fixed()

med_metab_df <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(filename, ms_run, station, depth, sla_corr),
            by="filename") %>%
  filter(ms_run=="MT") %>%
  group_by(depth, feature) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  ungroup() %>%
  group_by(filename, depth, station, sla_corr) %>%
  summarize(avg_amt=mean(norm_area)) %>%
  ungroup()
# Calculations for correlation between grand median metabolite and SLA
# med_metab_df %>%
#   group_by(station, depth, sla_corr) %>%
#   summarise(grand_avg_amt=mean(avg_amt)) %>%
#   group_by(depth) %>%
#   summarise(sim=cor(sla_corr, grand_avg_amt))
med_metab_sla_gp <- med_metab_df %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_boxplot(aes(x=station, y=avg_amt, color=sla_corr), linewidth=0.8) +
  facet_wrap(~depth, nrow=1) +
  scale_color_gradient2(low = scales::muted("blue"), high = scales::muted("red"),
                        mid="grey90", aesthetics = c("color", "fill"), guide = NULL) +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text = element_blank()) +
  labs(x="Station", y="Median metabolite z-score")

@@

*Figure 1: Plots of the metabolome across the eddy dipole, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. NMDS stress values (s) have been reported in the bottom left corner, while PERMANOVA R$^2$ and p-values are reported in the top left. SLA trends are visible in the DCM and deep samples, with dark blue circles consistently discriminating from the dark red circles along the first multidimensional axis. The bottom row of plots show the direction and magnitude of this effect by plotting the grand median of the z-score normalized metabolite peak areas in the stations taken across the eddy dipole with the upper whisker extent, median bar, and lower whisker extent corresponding to the values measured in three biological triplicates.*

We expected a large part of the SLA trends to be due to the elevated biomass in the cyclone DCM relative to the anticyclone and the inverse pattern in the 175 meter samples. While we were unable to quantify a majority of the molecular features in the dataset for direct comparison to biomass metrics, a subset of those were known metabolites that we quantified using standards (Figure 2).

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(9.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "10")
)
barplot_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=median(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x=NULL, y="Median nM\nmetabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  ggnewscale::new_scale_fill() +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), 
               data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  scale_color_identity() +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_blank(), 
        legend.margin = margin(t = 10,0,0,0), 
        legend.box.margin = margin(-10,-10,-10,-10),
        plot.margin = unit(numeric(4), "cm"),
        strip.text = element_text(face = "bold"),
        plot.background = element_rect(fill="white"))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)

@@

filled_file_metadata <- read_csv("../metadata/filled_file_metadata.csv") %>%
  filter(polarity=="pos") %>%
  select(-polarity)
clean_stans <- read_csv("../metadata/clean_stans.csv") %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(n_N=as.numeric(str_extract(formula, "(?<=N)\\d+"))) %>%
  mutate(n_N=case_when(
    !is.na(n_N)~n_N,
    str_detect(formula, "N")~1,
    TRUE~0
  ))

untarg_feats <- read_csv("../untargeted/all_feats.csv")
untarg_peaks <- read_csv("../untargeted/all_peaks.csv") %>%
  filter(str_detect(filename, "_Smp_"))
targ_peaks <- read_csv("../targeted/all_peaks.csv") %>%
    mutate(filename=paste0(filename, ".mzML"))

@@

Mesoscale eddies significantly alter open ocean environments such as those found in the subtropical gyres that cover a large fraction of the global ocean. Previous studies have explored eddy effects on biogeochemistry and microbial community composition but not on the molecular composition of particulate organic matter. This study reports the exact concentration of 67 metabolites and relative abundances for 640 molecular features to understand how mesoscale eddies impact the metabolome of the North Pacific Subtropical Gyre during two cruises in 2017 and 2018. We find that many metabolites track biomass trends, but biomarker metabolites like isethionic acid, homarine, and trigonelline linked to eukaryotic phytoplankton and enriched at the deep chlorophyll maximum of the cyclonic features, while degradation products such as arsenobetaine were enriched in anticyclones. By analyzing depth variability (accounting for 20-40% of metabolomic variability across ~150 meters) and eddy pumping (explaining 10-20% of variability across a sea level anomaly range of 40 centimeters), this analysis constrains the importance of mesoscale eddies in modeling the chemical composition of particulate matter in the largest biomes on the planet.

@@

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had previously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016].

Eddies are observed via satellite altimetry, which measures anomalies in sea surface height. Those that have a positive sea level anomaly (SLA) typically indicate regions of downwelling as water converges and are associated with clockwise rotation in the Northern Hemisphere. In contrast, a negative SLA usually corresponds to counterclockwise rotation in the Northern Hemisphere upwelling as deep water is lifted into the sunlit region of the ocean to replace water forced away from the eddy center [@McGillicuddy2016]. While mode-water eddies have been observed and are an exception to the conventions established above [@Sweeney2003; @McGillicuddy2007], here we focus on the more common cyclonic (negative SLA) and anticyclonic (positive SLA) mesoscale features near station ALOHA [@Barone2019].

The uplift of deep, nutrient-rich seawater into the euphotic zone disrupts microbial communities [@Rii2022]. Measurements of chlorophyll in cyclones reveal a shallower and more intense maximum [@Barone2022; @Cornec2021] as a result of eukaryotic phytoplankton growing in the increased nutrient concentrations while cyanobacterial biomass is reduced [@Hawco2021]. This growth of large eukaryotes corresponds to a net increase in biomass and primary productivity [@Benitez-Nelson2007], especially near the deep chlorophyll maximum [@Barone2022]. In contrast, anticyclones produce conditions favorable to nitrogen fixation by accumulating organisms such as Crocosphaera and Trichodesmium [@Fong2008; @Olson2015; @Dugenne2020; @Dugenne2023].

This shift in community structure and productivity should result in a corresponding shift in the composition of the particulate organic matter produced. Eukaryotic organisms have distinct chemical fingerprints from those of cyanobacteria and heterotrophic bacteria [@Heal2021; @Durham2022] and the relative contribution of these different taxa to total biomass should reflect their chemical composition. However, recent work from @Harke2021 showed that overall community function may be robust across eddy types at the surface while @Gleich2024 detected significant differences in protistan community composition and metabolic potential down to 250 meters, with heterotrophy-associated transcripts found at higher abundances in the cyclone. These results raise questions about the effect of mesoscale features on particulate organic matter composition.

# Results

We explored the impacts of isopycnal uplift and depression on the composition of particulate organic matter across pairs of mesoscale eddies of opposite polarity in the North Pacific Subtropical Gyre (NPSG). The first eddy pair was the focus of the July 2017 Microbial Ecology of the Surface Ocean-Simons Collaboration on Ocean Processes and Ecology (MESO-SCOPE, KM1709 aboard the R/V *Kilo Moana*) cruise and sampling consisted of two phases (Figure 1). In the first phase, a transect was taken from the north to the south consisting of 11 stations corresponding to the sampling sites in @Barone2022 and @Dugenne2023. Metabolomic samples of seawater from 15 meters, the deep chlorophyll maximum (DCM), and 175 meters were collected at each station via 0.2 µm filter. The next phase of the MESO-SCOPE cruise focused on Lagrangian sampling of the eddy centers, starting with the cyclone (SLA = -14 cm) before progressing to the anticyclone (SLA = 24 cm). There, metabolomics samples were taken from the DCM as well as 10 and 20 meters above and below it. The second eddy pair was targeted during the March 2018 Hawaiian Eddy Experiment (HEE, FK180310 aboard the R/V *Falkor*). During this followup cruise, a strong anticyclone (SLA = 21 cm) and a nearby cyclone (SLA = -13 cm) were targeted for a similar set of biogeochemical measurements described in @Dugenne2023 and @Gleich2024 (Figure 1). Metabolomics samples were taken from the center of each eddy as described above at both 25 meters and the DCM. We analyzed these three different datasets separately and discuss the findings from each in sequence below.

![](figures/MapForWill_v3.jpg)

*Figure 1: Sampling during MESO-SCOPE and the Hawaiian Eddy Experiment (HEE). Cruise bounds are shown in the large central map with Station ALOHA colored as a point in red near the Hawaiian Islands. Yellow stars denote sampling locations and station numbers relative to the sea level anomaly contours in the background during both the June 2017 MESO-SCOPE cruise (left) and the April 2018 HEE cruise (right).*

Metabolomics samples from all three sample sets were extracted via a modified Bligh & Dyer [@Boysen2018], with the aqueous fraction run on a HILIC column paired with a Q-Exactive HF mass spectrometer. Metabolites for which an authentic standard was available and run alongside were integrated in Skyline (HEE and MESO-SCOPE eddy centers) and MSDIAL (MESO-SCOPE eddy transect). Untargeted analysis was performed using the R package XCMS with each feature manually inspected for quality assurance.

## Metabolome variability across adjacent mesoscale eddies of opposing polarity during MESO-SCOPE explained by variations in sea level anomaly

The metabolome clearly differed between cyclonic and anticyclonic samples, though the magnitude of this difference varied by depth. The largest differences in particulate matter composition were detected at the DCM and 175 meter samples, as shown by the NMDS in Figure 2. At the DCM, approximately 16% of the variation in the metabolome could be explained by sea level anomaly (SLA) alone (PERMANOVA, p = 0.003). Additionally, SLA was strongly correlated with the first principal component (PC) of the metabolome  with a Pearson's *r* of .615 and 27% of the variance explained by PC1 (Supplemental figure 1A), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy at Station 12 were highly distinct and likely drove much of the variance explained. In the 175 meter samples, 12% of the variance was explained by SLA (PERMANOVA, p = 0.008) and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4, Supplemental figure 1A), though the first PC did not seem to have any visible pattern with metadata and appeared to largely capture variation between biological triplicates (Supplemental figure 1B). At 15 meters, SLA trends are less evident with a larger p-value (PERMANOVA, p=0.087) and a lower R$^2$ (0.09) (Figure 2).

```{r multivar_analysis object creation - nmds/pca/permanova}
set.seed(123)
multivar_analysis <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15 m)", "DCM  (110-120 m)", "Deep  (175 m)"))) %>%
  filter(ms_run=="MT") %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(feature, filename, norm_area) %>%
      group_by(feature) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      mutate(norm_area=scale(norm_area)[,1]) %>%
      pivot_wider(names_from = "feature", values_from = norm_area) %>%
      column_to_rownames("filename") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE) %>%
    vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(filename=rownames(metab_mat)) %>%
      left_join(filled_file_metadata %>% select(filename, sla_class), by="filename")
    vegan::adonis2(
      metab_mat~sla_class, data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(perm_r2=map_dbl(perm_output, pluck, "R2", 1)) %>%
  mutate(perm_pval=map_dbl(perm_output, pluck, "Pr(>F)", 1)) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(nmds_stress=paste("s =", round(nmds_stress, digits = 3))) %>%
  mutate(perm_r2=round(perm_r2, digits = 2)) %>%
  mutate(perm_pval=round(perm_pval, digits = 4)) %>%
  mutate(perm_label=paste0("R^2^ = ", perm_r2, "<br>p = ", perm_pval))
```

```{r figure nmds and median metabolite across eddy transect}
nmds_sla_gp <- multivar_analysis %>%
  mutate(nmds_points=map(nmds_output, function(nmds_output_i){
    nmds_output_i %>% pluck("points") %>% as.data.frame()
  })) %>%
  select(depth, nmds_points) %>%
  unnest_longer(nmds_points, indices_to = "filename") %>%
  unnest_wider(nmds_points) %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_corr)) %>%
  mutate(MDS2=ifelse(str_detect(depth,"Deep"), MDS2*-1, MDS2)) %>%
  ggplot() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, fill=sla_corr), color="black", size=4, pch=21) +
  geom_richtext(x=-Inf, y=Inf, aes(label=perm_label), hjust=0, vjust=1, fill="#FFFFFF33",
             data = multivar_analysis, label.r = unit(0, "in"), lineheight = .7) +
  geom_richtext(x=-Inf, y=-Inf, aes(label=nmds_stress), hjust=0, vjust=0, fill="#FFFFFF33",
             data = multivar_analysis, label.r = unit(0, "in")) +
  facet_wrap(~depth, nrow=1) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"),
  ) +
  labs(fill="Corr.\nSLA\n(cm)\n", x="NMDS axis 1", y="NMDS axis 2") +
  theme_bw() +
  theme(strip.text = element_text(size = 12)) +
  coord_fixed()

med_metab_df <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(filename, ms_run, station, depth, sla_corr),
            by="filename") %>%
  filter(ms_run=="MT") %>%
  group_by(depth, feature) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  ungroup() %>%
  group_by(filename, depth, station, sla_corr) %>%
  summarize(avg_amt=mean(norm_area)) %>%
  ungroup()
# Calculations for correlation between grand median metabolite and SLA
# med_metab_df %>%
#   group_by(station, depth, sla_corr) %>%
#   summarise(grand_avg_amt=mean(avg_amt)) %>%
#   group_by(depth) %>%
#   summarise(sim=cor(sla_corr, grand_avg_amt))
mean_med_metab_df <- med_metab_df %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  group_by(depth, station, sla_corr) %>%
  summarise(mean_val=mean(avg_amt))
med_metab_sla_gp <- med_metab_df %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  # geom_col(aes(x=station, y=mean_val, fill=sla_corr), color="black", 
  #            data=mean_med_metab_df) +
  geom_point(aes(x=station, y=avg_amt), color="black", size=2) +
  geom_point(aes(x=station, y=mean_val, fill=sla_corr), color="black",
             size=4, pch=21, data=mean_med_metab_df, alpha=0.7) +
  facet_wrap(~depth, nrow=1) +
  scale_color_gradient2(low = scales::muted("blue"), high = scales::muted("red"),
                        mid="grey90", aesthetics = c("color", "fill"), guide = NULL) +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text = element_blank()) +
  labs(x="Station", y="Median metabolite z-score")

# med_metab_sla_gp <- med_metab_df %>%
#   mutate(station=factor(station, levels=as.character(4:14))) %>%
#   mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
#   ggplot() +
#   geom_hline(yintercept = 0) +
#   geom_point(aes(x=station, y=avg_amt, fill=sla_corr), color="black", size=4, pch=21) +
#   facet_wrap(~depth, nrow=1) +
#   scale_color_gradient2(low = scales::muted("blue"), high = scales::muted("red"),
#                         mid="grey90", aesthetics = c("color", "fill"), guide = NULL) +
#   theme_bw() +
#   theme(strip.background = element_blank(), strip.text = element_blank()) +
#   labs(x="Station", y="Median metabolite z-score")

@@

*Figure 2: Distribution of metabolites in multivariate space across adjacent eddies of opposite polarity during the MESO-SCOPE transect, broken down by depth. Top panels depict non-metric multidimensional scaling (NMDS) plots with individual samples colored based on their corrected sea level anomaly. NMDS stress values (s) have been reported in the bottom left corner, while PERMANOVA R$^2$ and p-values are reported in the top left. SLA trends are visible in the DCM and 175 meter samples, with dark blue circles consistently discriminating from the dark red circles along the first multidimensional axis. Bottom panels depict the direction and magnitude of this effect by plotting the grand mean value of three biological triplicates in color and the raw values in black behind.*

To characterize the difference between eddies in a lower-dimensional space, we plotted an average metabolite peak area per station at each depth. We performed z-score normalization on each mass feature to give them all equal weight while preserving the variance between samples then calculated the median z-score for each sample, resulting in large positive values when metabolites are more abundant in a given sample than in the median sample (Figure 2). The average metabolite collected at the exact center of the cyclonic eddy (Station 12) had higher values than other samples at this same depth, indicating that most metabolites were more abundant at the cyclone's DCM than elsewhere in the transect. The 175 meter samples, on the other hand, showed the largest metabolite abundances in the anticyclone while the cyclone had consistently lower median z-scores. This method highlighted the negative correlation between the average metabolite and SLA at the DCM ($r_{DCM}= -.716$) while the 15 meter and 175 meter samples showed the opposite trend ($r_{15m} = .211$, $r_{175m} = .830$).

```{r perc increase at cyclone DCM center, eval=FALSE}
filled_file_metadata %>%
  filter(ms_run=="MT") %>%
  filter(depth=="DCM") %>%
  distinct(cruise, station, cast, depth, PC_um) %>%
  group_by(MT_DCM_cyc_center=ifelse(station==12, "St12", "Other")) %>%
  summarise(mean_val=mean(PC_um)) %>%
  pivot_wider(names_from=MT_DCM_cyc_center, values_from = mean_val) %>%
  mutate(perc_increase=St12/Other)

targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  filter(depth=="DCM") %>%
  group_by(filename, station, tripl, PC_um) %>%
  summarize(tot_nM=sum(nM)) %>%
  group_by(MT_DCM_cyc_center=ifelse(station==12, "St12", "Other")) %>%
  # group_by(MT_DCM_cyc_center=ifelse(station==12, "St12", "Other"), tripl) %>%
  summarise(mean_val=mean(tot_nM)) %>%
  pivot_wider(names_from=MT_DCM_cyc_center, values_from = mean_val) %>%
  mutate(perc_increase=St12/Other)
```

```{r PC_vs_totnM regression values, include=FALSE}
frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity, n_C, n_N)) %>%
  mutate(nM_C=n_C*nM, nM_N=n_N*nM) %>%
  group_by(filename) %>%
  summarize(total_nM_C=sum(nM_C), total_nM_N=sum(nM_N)) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(frac_C_in_metab=total_nM_C/1000/PC_um,
         frac_N_in_metab=total_nM_N/1000/PN_um) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  ungroup()

# Overall trend between PC and total nM metabolites (all depths)
broom::tidy(lm(total_nM_C~PC_um, data=no_outlier_frac_pc_in_metab_df))[2,]
with(no_outlier_frac_pc_in_metab_df, cor(PC_um, total_nM_C))
lme4::lmer(total_nM_C~PC_um + (1|depth), data=no_outlier_frac_pc_in_metab_df)

# By-depth trend between PC and total nM metabolites
no_outlier_frac_pc_in_metab_df %>%
  group_by(depth) %>%
  summarise(overall=broom::tidy(lm(total_nM_C~PC_um))[2,], 
            cor_vals=cor(PC_um, total_nM_C))
